---
title: "Fertility"
format: pdf
editor: visual
---

## Dataset tidying and merging

#### 1) Importing the dataset

```{r}
library(tidyverse)
library(corrplot)
library(binsreg)


Fertility <- read_csv("Fertility.csv")
View(Fertility)
fertility <- Fertility
```

#### 2) Tidying dataset

```{r}
fertility <- fertility %>% select(-c(TIPO_DATO15, `Flag Codes`, Flags, `Seleziona periodo`, ITTER107))
fertility<- fertility %>%  group_by(Territorio) %>% filter(TIME== 2021) %>%  distinct()

```

#### 3) Pivot wider

```{r}
fertility <- fertility %>% group_by(Territorio) %>% pivot_wider(names_from = c(`Tipo dato`), values_from = Value)  %>%  select(!TIME)

```

#### 4) Summary

```{r}

summary(fertility)

```

#### 

## Childcare

#### 1) Import dataset

```{r}
library(readr)
DCIS_SERVSOCEDU1_27112023145214126 <- read_csv("DCIS_SERVSOCEDU1_27112023145214126.csv")
Social_Services <- DCIS_SERVSOCEDU1_27112023145214126
View(Social_Services)
```

#### 2) Tidying: Social Services

```{r}
soc_serv <- Social_Services 
soc_serv<- soc_serv %>%  filter(`Tipo di servizio socio-educativo`!= "spazi gioco" & `Tipo dato`== 'posti autorizzati per 100 bambini di 0-2 anni' & `Settore del titolare`== 'totale' & `Tipo di servizio socio-educativo` != "tutte le voci")
soc_serv<- soc_serv %>% select(-c(TIPSERVSOC, SETTORE, `Settore del titolare`, `Seleziona periodo`, `Flag Codes`, Flags, TIPO_DATO_SERV_SOC, ITTER107))

soc_serv <- soc_serv %>% group_by(Territorio, TIME) %>% summarise("Total_places"= sum(Value))
soc_serv_years<- soc_serv %>% pivot_wider(names_from = c(TIME), values_from = Total_places)


 

```

#### 4) Childcare places availability mean 2017 to 2021

```{r}
soc_serv_mean <- soc_serv %>%  group_by (Territorio) %>% summarise (childcare_2017_2021_mean = mean(Total_places))
summary(soc_serv_years)
```

## Childcare & Fertility

```{r}
Childcare_Fertility <- inner_join(soc_serv_mean, fertility, by = join_by(Territorio))
Childcare_Fertility <- `Childcare_Fertility` %>% rename ('Total fertility index' =`tasso di fecondità totale`, 'Places in childcare' = childcare_2017_2021_mean, 'Average fathers age' = 'età media dei padri alla nascita del figlio', 'Average mothers age' = 'età media delle madri al parto', 'Territory' = 'Territorio' )

Childcare_Fertility <- Childcare_Fertility %>%  relocate(`Total fertility index`, .after = Territory)
summary(Childcare_Fertility)
```

```{r}
Childcare_Fertility <- Childcare_Fertility %>%  arrange(desc(`Total fertility index`))
```

## Omitted Variables

### A) Gender Occupation

```{r}
## Donne che lavorano

library(readr)
library(tidyverse)
library(ggplot2)
DCCV_TAXOCCU1_30112023101631613 <- read_csv("DCCV_TAXOCCU1_30112023101631613.csv")
View(DCCV_TAXOCCU1_30112023101631613)
'Employement Women' <- DCCV_TAXOCCU1_30112023101631613




```

#### 1) Filter

```{r}
library(dplyr)
`Employement Women`<- `Employement Women` %>%  filter(Gender == 'females' & TITOLO_STUDIO == 99 & TIME == 2022 & ETA1 == 'Y15-64')

```

#### 2) Select

```{r}

`Employement Women`<- `Employement Women` %>% select(c(Territory, Value))
`Employement Women`<- `Employement Women` %>% rename( 'Women employement rate' = Value)
summary(`Employement Women`)

```

#### 3) Merge

```{r}
Childcare_Fertility<- Childcare_Fertility %>%  left_join (`Employement Women`, by = join_by(Territory))
View(Childcare_Fertility)

#Is it omitted variable?

library(corrr)
cor(Childcare_Fertility_[,2], Childcare_Fertility_[,4]) #   0.2728453
cor(Childcare_Fertility_[,3], Childcare_Fertility_[,4]) # 0.9196262
```

### B) Income

```{r}
DCCV_REDNETFAMFONTERED_01122023102843294 <- read_csv("DCCV_REDNETFAMFONTERED_01122023102843294.csv")
Income <- DCCV_REDNETFAMFONTERED_01122023102843294
View(Income)

```

#### Tidying

```{r}
Income <- Income |>filter(`Data type` == 'annual average households income' , `Households main income source`== 'total', `Including or not including imputed`== 'not including imputed rents')
Income <- Income %>% select(Territory, Value)
Trentino <- as.data.frame(Income[8:9,2])
Trentino <- mean(Trentino$Value)

```

#### Join

```{r}
library(dplyr)
Childcare_Fertility_ <- Childcare_Fertility_ %>% left_join(Income) 
Childcare_Fertility_ <- Childcare_Fertility_ %>% rename('Household Income' = Value) 
#### Fill missing value
Childcare_Fertility_[1,6]= Trentino
```

## PCA: Variables

### A) Poverty

#### 1) Importing the Dataset

```{r}
library(tidyverse)

DCCV_AVQ_FAMIGLIE_01122023171754727 <- read_csv("Programming Lab/Project/Dataset usati/DCCV_AVQ_FAMIGLIE_01122023171754727.csv")
'Economic condition perceived' <- DCCV_AVQ_FAMIGLIE_01122023171754727
View(`Economic condition perceived`)
```

#### 2) Tidying Dataset

```{r}
`Economic condition perceived` <- `Economic condition perceived` %>% filter(TIME == 2021, Measure == 'per 100 households with the same characteristics')
`Economic condition perceived` <- `Economic condition perceived` %>% select(Territory, 'Data type', Value)
`Economic condition perceived` <- `Economic condition perceived` %>% pivot_wider(names_from = 'Data type', values_from = Value)


`Economic condition perceived` <- `Economic condition perceived` %>% mutate('Poor' = (`Economic condition perceived`[,8] + `Economic condition perceived`[,9])) 

`Economic condition perceived` <- `Economic condition perceived` %>% select(1,10)
summary(`Economic condition perceived`)


```

#### 3) Merge

```{r}
Childcare_Fertility_ <- Childcare_Fertility %>% left_join(`Economic condition perceived`) %>%  rename('Poor Conditions' =`Poor`  )
Childcare_Fertility_ <- Childcare_Fertility_ %>%  select(!4:5)



```

### B) Religion

```{r}
library(readr)
religion <- read_csv("religion.csv")
View(religion)
Religion <- religion
```

#### Tidying

```{r}

Religion <- Religion %>% filter(!`Data type`== 'never', 
Measure == 'per 100 people with the same characteristics',
TIME == 2021)
Religion <- Religion %>%  select(Territory, Value) 

```

#### Join

```{r}
Childcare_Fertility_ <- Childcare_Fertility_ %>%  left_join(Religion) %>% rename('Religion' = Value)


```

### C) Study years

Previous reallignement with the aid of excel was necessary given that the dataset misplaced some values when downloaded; the dataset already only contains the data regarding women.

```{r}
library(readr)
Popolazione_per_titolo_di_studio_e_regioni_IT1_52_1194_DF_DCCV_POPTIT1_UNT2020_2_1_0_2_ <- read_csv("Popolazione per titolo di studio e regioni (IT1,52_1194_DF_DCCV_POPTIT1_UNT2020_2,1.0) (2).csv")
'Study years' <- Popolazione_per_titolo_di_studio_e_regioni_IT1_52_1194_DF_DCCV_POPTIT1_UNT2020_2_1_0_2_

```

#### Tidying

```{r}
library(tidyverse)
library(dplyr)
`Study years` <- `Study years` %>% filter(EDU_LEV_HIGHEST != 99)

#Change the string values to year values

`Study years` <- `Study years`  %>% 
  mutate(years_study = case_when( EDU_LEV_HIGHEST == 3 ~ 5,
  EDU_LEV_HIGHEST == 4 ~ 8,
 EDU_LEV_HIGHEST == 5 ~ 11,
   EDU_LEV_HIGHEST == 6 ~ 13,
   EDU_LEV_HIGHEST == 11 ~ 16))

`Study years` <- `Study years`  %>% rename(observations=Osservazione)
`Study years` <- `Study years`  %>%   select(Territorio, years_study, observations)

```

#### Computing the regional study years average of women

```{r}

`Study years`<- `Study years` %>% 
  mutate(product= years_study * observations)

`Study years`<- `Study years` |>
  group_by(Territorio) |>
  mutate(totalyears_per_region=
           sum(product))

  
  `Study years` <- `Study years` |>
    group_by(Territorio) |>
    mutate(total_n_per_region=sum(observations))


  

```

```{r}
`Study years` <- `Study years` |>
  mutate('Regional study years average'= totalyears_per_region / total_n_per_region)

`Study years`<-`Study years` |>
  select(Territorio, 'Regional study years average')

`Study years`<-`Study years` |>
  distinct() %>%  rename(Territory = Territorio)

view(`Study years`)
```

#### Join

```{r}
Childcare_Fertility_ <- Childcare_Fertility_ %>%  left_join((`Study years`), join_by(Territory))
Childcare_Fertility_[7,8] = 10.467936

```

### D) Foreign residents

#### Importing dataset

```{r}
library(tidyverse)
library(dplyr)

foreign_residents<-read_csv("stranieri residenti.csv", show_col_types = FALSE)
foreign_residents <- stranieri_residenti
view(foreign_residents)
```

#### Tidying dataset

```{r}
foreign_residents<-foreign_residents|>
  filter(SEXISTAT1==9, ETA1=="TOTAL")|>
  select(-ITTER107, -SEXISTAT1, -Gender, -Age, -TIME, -`Select time`, -`Flag Codes`, -Flags, -ETA1,-TIPO_DATO15) %>% distinct()
```

#### Pivot wider

```{r}
foreign_residents<-foreign_residents|>
  group_by(Territory)|>
  pivot_wider(names_from = `Demographic data type`, values_from = Value)
```

#### Join

```{r}

Childcare_Fertility_ <- Childcare_Fertility_ %>% 
  left_join(foreign_residents)|> rename(foreign_residents = 'population on 1st January')
## Fill missing value
Childcare_Fertility_[1,9]= foreign_residents[32,2]+ foreign_residents[34,2]
  
```
